---
title: "Part4_Analyses_on_Harmonized_Data"
date: "2024-09-18"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
---

This file assumes that the code in `Part1_Harmonization_continuous.Rmd`, `Part2_Harmonization_ordinal.Rmd`, and `Part3_Results.Rmd` have been run. These files result in the `df.rds` file that will be read in within the current file. If those files have not been run yet, please run the files before executing the code below. 

# Install and load packages, prepare data.

```{r echo = TRUE, warning=FALSE, message = FALSE}
library(dplyr)
library(lavaan)
library(sirt)
library(mirt)
library(kableExtra)
library(here)
library(ggplot2)
library(umx)
# also install packages: haven, numDeriv
```

```{r echo = TRUE, message = FALSE, warning = FALSE}
score_df_cont <- readRDS("rds/score_df_continuous.rds")
score_df_ord <- readRDS("rds/score_df_ord.rds")
df <- cbind(score_df_cont, score_df_ord[,c("approx_ord", "approx_ord_SE", "approx_ord_rel", "approx_ord_ev")])
head(df, 2)
dim(df)
```

```{r}
# outcome variable: dropout
table(df$dropout)
round(table(df[df$sample == "ELS", ]$dropout) / 
        sum(table(df[df$sample == "ELS", ]$dropout)), 3)
round(table(df[df$sample == "HSLS", ]$dropout) / 
        sum(table(df[df$sample == "HSLS", ]$dropout)), 3)
```

# EAP Factor score regression (no reliability adjustment)

## Using EAP scores to predict dropout with probit regression

```{r}
# Warning: the code requires data to be sorted by `sample` before analyses
df$dropout <- as.numeric(df$dropout)
glm_approx_ord_int <- glm("dropout ~ approx_ord * sample", data = df, 
                      family = "binomial")
summary(glm_approx_ord_int)
# interaction not significant
glm_approx_ord_main <- glm("dropout ~ approx_ord + sample", data = df, 
                      family = "binomial")
summary(glm_approx_ord_main)
```

## Using `umx` with EAP scores (pooled analysis without unreliability adjustment)
  
```{r}
# create a dummy variable for the ELS sample (n_group - 1 dummy variables needed)
df$ELS <- 0
df$ELS <- ifelse(df$sample == "ELS", 1, 0)
```

```{r}
# Recode discrete variable to be used with OpenMx
df$dropout_f <- mxFactor(df$dropout, levels = c(0, 1))
```

```{r warning = FALSE, message = FALSE}
probitreg_umx_approx_ord <- umxRAM("
    dropout_f ~ approx_ord + ELS
    dropout_f ~ 0 * 1
    dropout_f ~~ 1 * dropout_f
  ",
  data = df,
  tryHard = "ordinal")
# plot(probitreg_umx_approx_ord) 
```

# Two-Stage Path Analysis with EAP scores 

Adjusting for measurement error
  
```{r warning = FALSE, message = FALSE}
# drop rows with missing reliability
df2 <- df[!is.na(df$approx_ord_rel), ]

# Reliability `approx_ord_rel` defined as 1 - approx_ord_SE^2 / psi_align_ord 
# (computed for each observation using the psi value for the correct group, see 
# `Part2_Harmonization_ordinal.Rmd` for details)

# Error variance `approx_ord_ev` defined as approx_ord_SE^2 * approx_ord_rel

probitreg_2spa_approx_ord <- umxRAM(
    "2spa",
    # Main effects of EAP scores and study indicators
    umxPath(c("eta", "ELS"), to = "dropout_f"),
    # Loading = reliability (as definition variable)
    umxPath("eta", to = "approx_ord", labels = "data.approx_ord_rel", free = FALSE),
    # Error variance (as definition variable)
    umxPath(var = "approx_ord", labels = "data.approx_ord_ev", free = FALSE),
    # Covariances of predictors
    umxPath(unique.pairs = c("eta", "ELS")),
    # Means of predictors
    umxPath(means = c("eta", "ELS")),
    # For model identification: Fix latent variate to be standard normal
    umxPath(v1m0 = "dropout_f"),
    data = df2,
    tryHard = "ordinal"
)
```



# Bartlett Factor score regression (no reliability adjustment)

## Using Bartlett scores to predict dropout with probit regression

Bartlett scores computed using the approximate invariance model assuming continuous data.
```{r}
glm_approx_cont_int <- glm("dropout ~ approx_cont * sample", data = df, 
                      family = "binomial")
summary(glm_approx_cont_int)
# interaction not significant
glm_approx_cont_main <- glm("dropout ~ approx_cont + sample", data = df, 
                      family = "binomial")
summary(glm_approx_cont_main)
```


Bartlett scores computed using the partial invariance model assuming continuous data.
```{r}
glm_partial_cont_int <- glm("dropout ~ partial_cont * sample", data = df, 
                      family = "binomial")
summary(glm_partial_cont_int)
# interaction not significant
glm_partial_cont_main <- glm("dropout ~ partial_cont + sample", data = df, 
                      family = "binomial")
summary(glm_partial_cont_main)
```

## Using `umx` with Bartlett scores (pooled analysis without unreliability adjustment)
  
Bartlett scores computed using the approximate invariance model assuming continuous data.
```{r warning = FALSE, message = FALSE}
probitreg_umx_approx_cont <- umxRAM("
    dropout_f ~ approx_cont + ELS
    dropout_f ~ 0 * 1
    dropout_f ~~ 1 * dropout_f
  ",
  data = df,
  tryHard = "yes")
# plot(probitreg_umx_approx_cont) 
```

Bartlett scores computed using the partial invariance model assuming continuous data.
```{r warning = FALSE, message = FALSE}
probitreg_umx_partial_cont <- umxRAM("
    dropout_f ~ partial_cont + ELS
    dropout_f ~ 0 * 1
    dropout_f ~~ 1 * dropout_f
  ",
  data = df,
  tryHard = "yes")
# plot(probitreg_umx_partial_cont) 
```



# Two-Stage Path Analysis with Bartlett scores 

Adjusting for measurement error
  
```{r warning = FALSE, message = FALSE}
# drop rows with missing reliability
df3 <- df[!is.na(df$approx_rel), ]

# Reliability set to 1 for Bartlett scores ("values = 1" below)
# Error variance `approx_ev` defined as approx_SE^2

probitreg_2spa_approx_cont <- umxRAM(
    "2spa",
    # Main effects of Bartlett scores and study indicators
    umxPath(c("eta", "ELS"), to = "dropout_f"),
    # Loading = reliability (as definition variable)
    umxPath("eta", to = "approx_cont", values = 1, free = FALSE),
    # Error variance (as definition variable)
    umxPath(var = "approx_cont", labels = "data.approx_ev", free = FALSE),
    # Covariances of predictors
    umxPath(unique.pairs = c("eta", "ELS")),
    # Means of predictors
    umxPath(means = c("eta", "ELS")),
    # For model identification: Fix latent variate to be standard normal
    umxPath(v1m0 = "dropout_f"),
    data = df3,
    tryHard = "yes"
)
```

```{r warning = FALSE, message = FALSE}
# drop rows with missing reliability
df4 <- df[!is.na(df$partial_rel), ]

# Reliability set to 1 for Bartlett scores ("values = 1" below)
# Error variance `partial_ev` defined as partial_SE^2

probitreg_2spa_partial_cont <- umxRAM(
    "2spa",
    # Main effects of Bartlett scores and study indicators
    umxPath(c("eta", "ELS"), to = "dropout_f"),
    # Loading = reliability (as definition variable)
    umxPath("eta", to = "partial_cont", values = 1, free = FALSE),
    # Error variance (as definition variable)
    umxPath(var = "partial_cont", labels = "data.partial_ev", free = FALSE),
    # Covariances of predictors
    umxPath(unique.pairs = c("eta", "ELS")),
    # Means of predictors
    umxPath(means = c("eta", "ELS")),
    # For model identification: Fix latent variate to be standard normal
    umxPath(v1m0 = "dropout_f"),
    data = df4,
    tryHard = "yes"
)
```


# Sensitivity Analysis

Using sum/mean scores

```{r}
# Scatterplot
plot(x = df$mean_score, y = df$partial_cont, xlab = "Mean scores", 
     ylab = "Bartlett scores (partial invariance, continuous)", cex.lab = 0.8)
cor(df$mean_score, df$partial_cont, use = "complete")
plot(x = df$mean_score, y = df$approx_ord, xlab = "Mean scores", 
     ylab = "Bartlett scores (approx. invariance, continuous)", cex.lab = 0.8)
cor(df$mean_score, df$approx_cont, use = "complete")
plot(x = df$mean_score, y = df$approx_ord, xlab = "Mean scores",
     ylab = "EAP scores (approx. invariance, ordinal)", cex.lab = 0.8)
cor(df$mean_score, df$approx_ord, use = "complete")
```

### Probit regression with mean scores

Note that the scale of the predictors are different, so the coefficients are not comparable.

```{r}
# GLM (probit regression); interaction not significant
glm_meanscore <- glm("dropout ~ mean_score * sample", data = df, 
                      family = "binomial")
summary(glm_meanscore)
# interaction not significant
glm_meanscore_main <- glm("dropout ~ mean_score + sample", data = df, 
                      family = "binomial")
summary(glm_meanscore_main)
```

