---
title: "Part3_Results"
author: "Meltem Ozcan"
date: "2024-09-12"
output: pdf_document
---

This file assumes that the code in `Part1_Harmonization_continuous.Rmd` and `Part2_Harmonization_ordinal.Rmd` have been run. These two files result in the rds files that will be read in within the current file. If those files have not been run yet, please run the files to prepare the relevant rds files. 

# Install and load packages, prepare data.

```{r echo = TRUE, warning=FALSE, message = FALSE}
library(dplyr)
library(lavaan)
library(sirt)
library(mirt)
library(kableExtra)
library(here)
library(ggplot2)
# also install packages: haven, numDeriv
```

```{r echo = TRUE, message = FALSE, warning = FALSE}
dat <- readRDS("rds/dat.rds")

est_partial <- readRDS("rds/est_partial.rds")
est_align <- readRDS("rds/est_align.rds")
est_partial_ord <- readRDS("rds/est_partial_ord.rds")
est_align_ord <- readRDS("rds/est_align_ord.rds")
est_align_ord_th <- readRDS("rds/est_align_ord_th.rds")
est_align_ord_lam <- readRDS("rds/est_align_ord_lam.rds")
fit_partial <- readRDS("rds/fit_partial.rds")
fit_align <- readRDS("rds/fit_align.rds")
mod_aligned <- readRDS("rds/mod_aligned.rds")

score_df_cont <- readRDS("rds/score_df_continuous.rds")
score_df_ord <- readRDS("rds/score_df_ord.rds")
df <- cbind(score_df_cont, score_df_ord[,c("approx_ord", "approx_ord_SE", "approx_ord_rel", "approx_ord_ev")])
head(df, 2)
saveRDS(df, "rds/df.rds")
m_items <- paste0("i", 1:5) 
```

Also source the following R file to repeat the analyses up to this point on the dataset without any NAs for later sensitivity analyses.

```{r}
if (!all(file.exists("rds/score_df_noNAs.rds",
                     "rds/fit_partial_noNAs.rds",
                     "rds/fit_align_noNAs.rds",
                     "rds/est_align_ord_noNAs.rds"))) {
  source("code/repeatAnalysesNoNAs.R")
} else {
  complete_cases <- complete.cases(dat[, m_items]) & dat$sample == "ELS" |
    complete.cases(dat[, m_items[-3]]) & dat$sample == "HSLS"
  d_noNAs <- dat[complete_cases, ]
  fit_partial_noNAs <- readRDS("rds/fit_partial_noNAs.rds")
  fit_align_noNAs <- readRDS("rds/fit_align_noNAs.rds")
  est_align_ord_noNAs <- readRDS("rds/est_align_ord_noNAs.rds")

  score_df_noNAs <- readRDS("rds/score_df_noNAs.rds")
}
```



```{r echo=FALSE}
models <- rbind("N = 30,749" = c("M_B1", "M1", "M2", "-", "M3"), 
                "N = 29,202" = c("M_B2", "M4", "M5", "M6", "M7"))
colnames(models) <- c("Mean Score", rep(c("Partial", "Approximate"), 2))

kbl(models, booktabs = T, align = "c", table.attr = "style='width:60%;'",
    caption = "Models by approach and dataset", longtable = TRUE) %>%
  kable_classic() %>% 
  add_header_above(c(" " = 2, "Continuous" = 2, "Ordinal" = 2))
```
```{r echo = TRUE}
source('code/table_helper_functions.R')
```

```{r echo = FALSE}
build_loadings_tab()
```

```{r echo = FALSE}
build_intercepts_tab()
```

\newpage

```{r echo = FALSE}
build_thresholds_tab()
```

```{r echo = FALSE}
build_latent_tab()
```

## Score distributions

```{r echo = FALSE}
build_mn_sd_tab(df, score_df_noNAs)
```

## Score correlations

```{r}
# compute matrix of correlations
cor_30749 <- cor(df[, c("mean_score", "partial_cont", "approx_cont", "approx_ord")], use = "complete.obs")
```

```{r echo=FALSE}
cor_30749 <- round(cor_30749, 3)
cor_30749_add_row <- rbind(cor_30749[1:3,], rep("-",4), cor_30749[4,])
cor_30749_add_col <- cbind(cor_30749_add_row[,1:3], rep("-",5), cor_30749_add_row[,4])
colnames(cor_30749_add_col) <- rownames(cor_30749_add_col) <- as.character(models[1,])

cor_30749_add_col %>% 
  kbl(booktabs = T, align = "c", table.attr = "style='width:80%;'",
      caption = "Correlation of FS (N = 30,740)") %>%
  add_header_above(c(" ", "Mean scores" = 1,  "Partial" = 1, "Approximate" = 1,
                     "Partial" = 1, "Approximate" = 1)) %>%
  add_header_above(c(" ", " " = 1,  "Continuous" = 2, "Ordinal" = 2)) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE)

# repeat for the dataset without NAs
cor_29202 <- cor(as.matrix(apply(score_df_noNAs[, c("mean_score",
                                                  "partial_cont", 
                                                  "approx_cont",
                                                  "partial_ord",
                                                  "approx_ord")], 
                                 as.numeric, MARGIN = 2)))
colnames(cor_29202) <- rownames(cor_29202) <- as.character(models[2,])

# to match the apperance of the previous table
cor_29202_mat <- cor_29202 %>% round(3) %>% apply(MARGIN = 2, as.character) %>%
  as.data.frame()
rownames(cor_29202_mat) <- colnames(cor_29202_mat)
cor_29202_mat %>%
  kbl(booktabs = T, align = "c", table.attr = "style='width:80%;'",
      caption = "Correlation of FS on dataset with no NAs (N = 29,202)") %>%
  add_header_above(c(" ", "Mean scores" = 1,  "Partial" = 1, "Approximate" = 1,
                     "Partial" = 1, "Approximate" = 1)) %>%
  add_header_above(c(" ", " " = 1,  "Continuous" = 2, "Ordinal" = 2)) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE)
```

```{r echo = FALSE, warning=FALSE, message = FALSE}
source('code/scatterplots.R')
```


## Reliability

```{r}
mnsd(df$partial_rel)
mnsd(df$approx_rel)
mnsd(df$approx_ord_rel)
```

```{r echo = FALSE}
tab_rel <- round(
  cbind(mnsd(df[df$sample=="ELS",]$partial_rel),
       mnsd(df[df$sample=="HSLS",]$partial_rel),
       mnsd(df[df$sample=="ELS",]$approx_rel),
       mnsd(df[df$sample=="HSLS",]$approx_rel),
       mnsd(df[df$sample=="ELS",]$approx_ord_rel),
       mnsd(df[df$sample=="HSLS",]$approx_ord_rel)), 4)
tab_rel <- cbind(tab_rel[,1:4], rep("-", 2), rep("-", 2), tab_rel[,5:6])

rownames(tab_rel) <- c("Mean", "SD")
kbl(tab_rel, booktabs = T, align = "c", table.attr = "style='width:70%;'",
      caption = "Mean and SD of FS Reliabilities (N = 30,740)") %>%
  add_header_above(c(" ", rep(c("ELS", "HSLS"), 4))) %>% 
  add_header_above(c(" "= 1, "Partial" = 2, "Approximate" = 2, "Partial" = 2,
                     "Approximate" = 2)) %>% 
  add_header_above(c(" "= 1, "Continuous" = 4, "Ordinal" = 4))
```

```{r echo = FALSE}
par(las=2)
hist(df$partial_rel, col = "lightblue", 
     main = "Reliability of Bartlett factor scores (partial inv.; cont.)", 
     xlab = "Reliability", cex = 0.7,  cex.axis = 0.7, ylim = c(0,30000))
hist(df$approx_rel, col = "lightblue3", 
     main = "Reliability of Bartlett factor scores (approx. inv,; cont.)", 
     xlab = "Reliability", cex = 0.7,  cex.axis = 0.7, ylim = c(0,30000))
hist(df$approx_ord_rel, col = "lightblue4",
     main = "Reliability of EAP factor scores (approx. inv.; ordinal)", 
     xlab = "Reliability", cex = 0.7,  cex.axis = 0.7, ylim = c(0,30000))
```

```{r}
rel_30749 <- cor(as.matrix(apply(df[ ,c("partial_rel",
                                               "approx_rel", 
                                               "approx_ord_rel")], 
                                 as.numeric, MARGIN = 2)),
                 use = "complete.obs")
rel_30749 <- round(rel_30749, 4)
rel_30749_add_row <- rbind(rel_30749[1:2,], rep("-",3) , rel_30749[3,])
rel_30749_add_col <- cbind(rel_30749_add_row[,1:2], rep("-",4) , rel_30749_add_row[,3])
colnames(rel_30749_add_col) <- rownames(rel_30749_add_col) <- as.character(models[1,])[2:5]
```

```{r echo = FALSE}
rel_30749_add_col %>% 
  kbl(booktabs = T, align = "c", table.attr = "style='width:80%;'",
      caption = "Correlation of FS Reliabilities (N = 30,740)") %>%
  add_header_above(c(" ",  "Partial" = 1, "Approximate" = 1,
                     "Partial" = 1, "Approximate" = 1)) %>%
  add_header_above(c( " " = 1,  "Continuous" = 2, "Ordinal" = 2)) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE)
```

```{r echo = FALSE, message=FALSE, warning = FALSE}
# run source('code/scatterplots.R') first if not already run 
scatter_p(df, df$partial_rel, df$approx_rel,
          xlab = "Bartlett FS reliability (partial, continuous)",
          ylab = "Bartlett FS reliability (approximate, continuous)",
          "Correlation of reliability of factor scores computed using
MI modeling or alignment")

scatter_p(df, df$approx_rel, df$approx_ord_rel,
           xlab = "Bartlett FS reliability (approximate, continuous)",
           ylab = "EAP FS reliability (approximate, ordinal)",
          "Correlation of reliability of factor scores computed using
continuous or ordinal alignment")
```


## Dmacs effect sizes of noninvariance

We use the R package `pinsearch` to compute the Dmacs effect size of noninvariance for each approach. We use the `pin_effsize()` function and `dmacs_ordered()` for the ordinal case. Note that if there are more than two groups, `pin_effsize()` computes the Fmacs effect size when data are continuous, and `fmacs_ordered()` should replace `dmacs_ordered()` in the ordinal case.

### Continuous case

```{r echo = TRUE}
dmacs_partial <- pinsearch::pin_effsize(fit_partial)
dmacs_align <- pinsearch::pin_effsize(fit_align)
```
### Ordinal case
```{r}
# indicate which thresholds belong to which item
est_align_ord_th_sub <- est_align_ord_th[, -c(3, 8, 13)]
colnames(est_align_ord_th_sub) <- rep(c(1, 2, 4, 5), 3)

# compute pooled SD for dmacs_ordered()
vars <- apply(dat[m_items], MARGIN = 2,
              FUN = \(x) tapply(x, INDEX = dat$sample, FUN = var, na.rm = TRUE)
              )
(wgt_mat <- as.matrix(rbind(
  colSums(!is.na(dat[dat$sample=="ELS", m_items])),
  colSums(!is.na(dat[dat$sample=="HSLS", m_items]))
  )))
item_n <- ifelse(wgt_mat > 0, wgt_mat, NA)
pooled_sd <- sqrt(colSums(vars * (item_n - 1), na.rm = TRUE) /
                    colSums(item_n - 1, na.rm = TRUE)
                  )

dmacs_align_ord <- pinsearch::dmacs_ordered(
  thresholds = as.matrix(est_align_ord_th_sub),
  loadings = as.matrix(est_align_ord_lam)[, -3],
  link = "logit",
  pooled_item_sd = pooled_sd[-3]
)
```

```{r}
# compute dmacs effect size for ordinal MI modeling approach:
# first recompute pooled SD since rows with NAs had to be dropped in the ordinal 
# MI modeling approach
## note that now we're using d_noNAs as out dataset. 



vars_noNAs <- apply(d_noNAs[m_items], MARGIN = 2,
                    FUN = \(x) tapply(x, INDEX = d_noNAs$sample, 
                                      FUN = var, na.rm = TRUE)
              )
(wgt_mat_noNAs <- as.matrix(rbind(
  colSums(!is.na(d_noNAs[d_noNAs$sample=="ELS", m_items])),
  colSums(!is.na(d_noNAs[d_noNAs$sample=="HSLS", m_items]))
)))
item_n_noNAs <- ifelse(wgt_mat_noNAs > 0, wgt_mat_noNAs, NA)
pooled_sd_noNAs <- sqrt(colSums(vars_noNAs * (item_n_noNAs - 1), na.rm = TRUE) /
                          colSums(item_n_noNAs - 1, na.rm = TRUE)
                        ) 
# indicate which thresholds belong to which item
est_partial_ord_th_sub <- as.matrix(rbind(t(est_partial_ord$ELS$tau)[-c(7:9)],
                                          t(est_partial_ord$HSLS$tau)))
colnames(est_partial_ord_th_sub) <- c(rep(1, 3), rep(2, 3), rep(4, 3), rep(5, 3))
est_partial_ord_lam <- as.matrix(rbind(t(est_partial_ord$ELS$lambda)[-3],
                                       t(est_partial_ord$HSLS$lambda)))

dmacs_partial_ord <- pinsearch::dmacs_ordered(
  thresholds = est_partial_ord_th_sub,
  loadings = est_partial_ord_lam,
  link = "logit",
  pooled_item_sd = pooled_sd_noNAs[-3]
)

```

```{r include = FALSE}
# repeat for the dataset with no NAs
dmacs_partial_noNAs <- pinsearch::pin_effsize(fit_partial_noNAs)
dmacs_align_noNAs <- pinsearch::pin_effsize(fit_align_noNAs)

# ordinal alignment 
est_align_ord_th_noNAs <- cbind(est_align_ord_noNAs[1:2,c("d1", "d2", "d3")],
                                est_align_ord_noNAs[3:4,c("d1", "d2", "d3")],
                                est_align_ord_noNAs[5:6,c("d1", "d2", "d3")],
                                est_align_ord_noNAs[7:8,c("d1", "d2", "d3")], 
                                est_align_ord_noNAs[9:10,c("d1", "d2", "d3")])
est_align_ord_lam_noNAs <- cbind(est_align_ord_noNAs[1:2,c("a1")],
                                 est_align_ord_noNAs[3:4,c("a1")],
                                 est_align_ord_noNAs[5:6,c("a1")],
                                 est_align_ord_noNAs[7:8,c("a1")], 
                                 est_align_ord_noNAs[9:10,c("a1")])
# indicate which thresholds belong to which item
est_align_ord_th_sub_noNAs <- est_align_ord_th_noNAs[, -c(7:9)]
colnames(est_align_ord_th_sub_noNAs) <- c(rep(1, 3), rep(2, 3), rep(4, 3), rep(5, 3))
dmacs_align_ord_noNAs <- pinsearch::dmacs_ordered(
  thresholds = as.matrix(est_align_ord_th_sub_noNAs),
  loadings = as.matrix(est_align_ord_lam_noNAs)[, -3],
  link = "logit",
  pooled_item_sd = pooled_sd_noNAs[-3]
)
```


```{r echo = FALSE}
eff_size_tab <- t(rbind(c(round(dmacs_partial, 3), "-", "-"), 
                        round(dmacs_align, 3),
                        rep("-", 4),
                        round(dmacs_align_ord, 3)))

colnames(eff_size_tab) <- c("Partial (M1)", "Approximate (M2)",
                            "        -       ", "Approximate (M3)")
rownames(eff_size_tab) <- c("i1", "i2", "i4", "i5")

kbl(eff_size_tab, booktabs = T, align = "c", table.attr = "style='width:70%;'",
      caption = "Dmacs effect sizes of noninvariance by approach (N = 30,749)", 
    longtable = TRUE, digits = 2) %>%
    kable_classic() %>% 
    add_header_above(c("  " = 1, "Continuous" = 2, "Ordinal" = 2))
```

```{r echo = FALSE}
eff_size_tab_noNAs <- t(rbind(c(round(dmacs_partial_noNAs, 3), "-", "-"), 
                        round(dmacs_align_noNAs, 3),
                        round(dmacs_partial_ord, 3),
                        round(dmacs_align_ord_noNAs, 3)))

colnames(eff_size_tab_noNAs) <- c("Partial (M4)", "Approximate (M5)",
                            "Partial (M6)", "Approximate (M7)")
rownames(eff_size_tab_noNAs) <- c("i1", "i2", "i4", "i5")

kbl(eff_size_tab_noNAs, booktabs = T, align = "c", table.attr = "style='width:70%;'",
      caption = "Dmacs effect sizes of noninvariance by approach (N = 29,202)", 
    longtable = TRUE, digits = 2) %>%
    kable_classic() %>% 
    add_header_above(c("  " = 1, "Continuous" = 2, "Ordinal" = 2))
```


